How fluid-mechanical erosion creates anisotropic porous media

### 2022-12-13 docs/PhysicsPaper/StokesErosion.tex
NM

Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows~\cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows~\cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plane of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

Our method merges highly-efficient and highly-accurate boundary-integral equation (BIE) methods~\cite{baker1986boundary, moore2007evaluation, gray2019boundary} with stable interface evolution methods~\cite{hou-low-she1994, Moore2013} to simulate the erosion of dense suspensions of solid bodies in \edit{two-dimensional Stokes flows. We employ the Stokes equations because it is the regime relevant for groundwater applications~\cite{quaife2018boundary, chiu2020viscous}}. Originally inspired by related work in the high-Reynolds-number regime~\cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in~\cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies~\cite{chiu2020viscous}. Though the governing Stokes equations are linear, the nonlinear feedback between evolving microstructure and flow gives rise to highly complex and anisotropic configurations.

@@

\subsection{\edit{Boundary integral equation formulation}}
\label{sec:bie}
To accurately and efficiently solve the Stokes equations~\eqref{eqn:StokesEq}, we reformulate the system as a boundary integral equation (BIE). A BIE formulation has the advantage that all the unknowns are on the one-dimensional boundaries of the domain. That is, only the boundary of the complex geometry must be discretized which we do with a spectrally accurate Fourier basis. Applying the same approach as our previous works~\cite{chiu2020viscous, quaife2018boundary}, we represent the velocity as the sum of a double-layer potential and a combination of Stokeslets and rotlets~\cite{pow-mir1987}
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + \sum_{\ell=1}^{M}\left(
    S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
    \quad \xx \in \Omega,
  \label{eqn:velocity}
\end{align}
where 
\begin{align}
  \DDD[\eeta](\xx) = \frac{1}{\pi}\int_{\bd\Omega} 
  \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
  \eeta(\yy) \, ds_\yy,
  \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$. Note that $\bd\Omega$ includes both the eroding bodies and the outer boundary. The Stokeslets and the rotlets are
\begin{align}
  S[\lambda_\ell](\xx) = \frac{1}{4\pi} \left(-\log \rho_\ell +
  \frac{\rr_\ell \otimes \rr_\ell}{\rho_\ell^2} \right) \llambda_\ell,
  \quad
  R[\xi_\ell](\xx) = \frac{\rr_\ell^\perp}{\rho_\ell^2}\xi_\ell,
\end{align}
respectively, where $\rr_\ell = \xx - \cc_\ell$, \edit{$\rr_\ell^\perp =
(r_2,-r_1)$,} $\rho_\ell = \|\rr_\ell\|$, and $\cc_\ell$ is a point inside body $\ell$. If the density function, Stokeslets, and rotlets satisfy the second-kind boundary integral equation
\begin{subequations}
  \label{eqn:FredBIE}
  \begin{alignat}{3}
    \UU(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      \quad &&\xx \in \Gamma, \\
    \mathbf{0} &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      &&\xx \in \gamma_\ell, \: \ell = 1,\ldots,M, \\
    \llambda_\ell &= \int_{\gamma_\ell} \eeta(\yy) \, ds_\yy, 
      &&\ell = 1,\ldots,M, \\
    \xi_\ell &= \int_{\gamma_\ell} (\yy - \cc_\ell)^\perp \cdot 
      \eeta(\yy) \, ds_\yy, &&\ell = 1,\ldots,M,
  \end{alignat}
\end{subequations}
then the representation~\eqref{eqn:velocity} satisfies the Stokes
equations with the required boundary conditions~\eqref{eqn:StokesEq}. We
solve~\eqref{eqn:FredBIE} by discretizing $\gamma_\ell$ and $\Gamma$ at
equispaced collocation points, and then replacing the integrals with
quadrature rules. This results in a linear system with a
mesh-independent condition number, and it is solved iteratively with
GMRES. Instead of evaluating the Stokes double-layer
potential~\eqref{eqn:velocityDLP} as a contour integral in $\RR^2$, we
convert the integral to a sum of contour integrals around Jordan curves
in $\CC$. \edit{This conversion allows us to use quadrature formulas
whose errors are uniformly bounded independent of the proximity of the
bodies. Details of the conversion and the quadrature formulas are in
Appendix~\ref{appendix:A}.}

>>>>

Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows~\cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows~\cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plane of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

Our method merges \edit{spectrally}-accurate boundary-integral equation (BIE) methods~\cite{baker1986boundary, moore2007evaluation, gray2019boundary} with stable interface evolution methods~\cite{hou-low-she1994, Moore2013} to simulate the erosion of dense suspensions of solid bodies in \edit{two-dimensional Stokes flows. We employ the Stokes system because it is the regime relevant for groundwater applications~\cite{quaife2018boundary, chiu2020viscous}}. Originally inspired by related work in the high-Reynolds-number regime~\cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in~\cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies~\cite{chiu2020viscous}. Though the governing Stokes equations are linear, the nonlinear feedback between evolving microstructure and flow gives rise to highly complex and anisotropic configurations.

@@

\subsection{\edit{Boundary integral equation formulation}}
\label{sec:bie}
To \edit{numerically} solve the Stokes equations~\eqref{eqn:StokesEq}, we reformulate the system as a boundary integral equation (BIE). A BIE formulation has the advantage that all the unknowns are on the one-dimensional boundaries of the domain. That is, only the boundary of the complex geometry must be discretized which we do with a spectrally accurate Fourier basis. Applying the same approach as our previous works~\cite{chiu2020viscous, quaife2018boundary}, we represent the velocity as the sum of a double-layer potential and a combination of Stokeslets and rotlets~\cite{pow-mir1987}
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + \sum_{\ell=1}^{M}\left(
    S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
    \quad \xx \in \Omega,
  \label{eqn:velocity}
\end{align}
where 
\begin{align}
  \DDD[\eeta](\xx) = \frac{1}{\pi}\int_{\bd\Omega} 
  \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
  \eeta(\yy) \, ds_\yy,
  \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$. Note that $\bd\Omega$ includes both the eroding bodies and the outer boundary. The Stokeslets and the rotlets are
\begin{align}
  S[\lambda_\ell](\xx) = \frac{1}{4\pi} \left(-\log \rho_\ell +
  \frac{\rr_\ell \otimes \rr_\ell}{\rho_\ell^2} \right) \llambda_\ell,
  \quad
  R[\xi_\ell](\xx) = \frac{\rr_\ell^\perp}{\rho_\ell^2}\xi_\ell,
\end{align}
respectively, where $\rr_\ell = \xx - \cc_\ell$, \edit{$\rr_\ell^\perp =
(r_2,-r_1)$,} $\rho_\ell = \|\rr_\ell\|$, and $\cc_\ell$ is a point inside body $\ell$. If the density function, Stokeslets, and rotlets satisfy the second-kind boundary integral equation
\begin{subequations}
  \label{eqn:FredBIE}
  \begin{alignat}{3}
    \UU(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      \quad &&\xx \in \Gamma, \\
    \mathbf{0} &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      &&\xx \in \gamma_\ell, \: \ell = 1,\ldots,M, \\
    \llambda_\ell &= \int_{\gamma_\ell} \eeta(\yy) \, ds_\yy, 
      &&\ell = 1,\ldots,M, \\
    \xi_\ell &= \int_{\gamma_\ell} (\yy - \cc_\ell)^\perp \cdot 
      \eeta(\yy) \, ds_\yy, &&\ell = 1,\ldots,M,
  \end{alignat}
\end{subequations}
then the representation~\eqref{eqn:velocity} satisfies the Stokes
equations with the required boundary conditions~\eqref{eqn:StokesEq}. We
solve~\eqref{eqn:FredBIE} by discretizing $\gamma_\ell$ and $\Gamma$ at
equispaced collocation points, and then replacing the integrals with
quadrature rules. This results in a linear system with a
mesh-independent condition number, and it is solved iteratively with
GMRES. Instead of evaluating the Stokes double-layer
potential~\eqref{eqn:velocityDLP} as a contour integral in $\RR^2$, we
convert the integral to a sum of contour integrals around Jordan curves
in $\CC$. \edit{This conversion allows us to use quadrature formulas
whose errors are uniformly bounded independent of the proximity of the
bodies. Details of the conversion and the quadrature formulas are in
Appendix~\ref{appendix:A}.}
